Cyclone: Fault Tolerance Middleware for NVM Clusters

### 2016-03-31 paper/paper.tex
Additions to the paper.
Updates to paper.

This paper presents Cyclone - fault tolerance middleware that exposes a
replicated and durable heap. This heap can be directly manipulated using loads and
stores. Cyclone transparently provides both failure atomicity as well as
strongly consistent replication thereby giving strong guarantees to programmers
with minimal effort on their part. Under the hood Cyclone decouples modification
of local persistent memory from replication to maximize performance. In
addition, unlike traditional systems that provide strong replication, Cyclone
does not take checkpoints since NVM is directly manipulated by programmers and
therefore up to date till the last executed transaction. Instead, Cyclone
provides a novel NVM copy mechanism to bring up new nodes added to the cluster.
\end{abstract}

@@

\begin{figure}
{\scriptsize
\begin{verbatim}

@@

\section{Failure Atomicity}
Programmers typically find it hard to reason about failures when writing
code. This is a problem when directly manipulating data structures in persistent
memory. Depending on the specifics of the system architecture, updates to
persistent memory can be persisted out of order thereby leaving  the data
structure in an inconsistent state. To illustrate this consider a failure during
execution of the code to add a node to a linked list in
Figure~\ref{fig:example}. If either of the two updates to the next fields of the
previous or new node fail to be persisted, the linked list is left in a
disconnected state with one or more connected nodes inaccessible.

@@

\begin{figure}
{\scriptsize
\begin{verbatim}
TX_BEGIN()
{
  Initialize undo log
}

@@

A crucial feature of Cyclone is that it abstracts the details of the persistent
mechanism away from programmers. This allows the same code to be deployed on
clusters with a persistent storage medium such as 3D XPoint or on clusters with
battery backed NVDIMMs for example. In the former case, explicit {\tt PCOMMIT}
instructions are necessary for persistence [ref PMFS paper], while in the latter
case no such instructions are necessary as on failure the CPU caches and DRAM
are flushed to stable storage under battery (or capacitor) power. Even in the
NVDIMM case however, the undo log is necessary to protect against OS or
application crashes to ensure that the persistent memory is always left in a
consistent state.

\section{Replication}
Cyclone replicates the state of the NVM heap across multiple machines in a
cluster. Cyclone provides strong consistency - a majority of machines in the
quorum of replicas always agrees on the current state of the NVM heap. Unlike
traditional approaches to strong replication however, Cyclone is designed to
decouple replication from execution of user code modifying NVM. It does so by
encapsulating changes in to the NVM heap in RPC calls driven by client requests.
We chose this model since it closely reflects the use cases most suitable for
NVM-based clusters.

Cyclone replicates state \emph{by executing the same sequence of RPC calls on
  all members of the quorum}. Unlike traditional replication therefore Cyclone
does not replicate logs of changes to apply to replicas. Instead it replicates
the RPC call and its arguments, executing it on every replica. By executing the
same sequence of RPC calls on all machines we end up with the same
state. Cyclone uses RAFT under the hood for consensus on the contents of the log
of RPC calls - although any other similar state machine replication protocol
such as Paxos is also suitable. Cyclone provides the means to achieve exactly
once semantics to clients by numbering RPC calls and remembering the results of
the last executed RPC call.

@@

\subsection{Asynchronous Replication}
By default, Cyclone decouples execution of the RPC call on any replica from its
replication. This is particularly beneficial to performance since it allows
overlapping the update of non-volatile state with the network I/O associated
with replication. Persisting RAFT log entries, making the upcall to the
application and executing the RPC call in a failure atomic transaction takes a
significant amount of time and therefore it makes sense to overlap this with the
round-trip delay on the network for replication. This is possible in Cyclone and
not other traditional replication systems since we use failure atomic
transactions that cna be rolled back. Replication can fail due to various
reasons such as failure of nodes, a network partition or a new leader getting
elected in the underlying replication protocol (RAFT). Since failure atomic
transactions can be rolled back, we are able to speculatively apply the changes
to NVM without waiting for replication to finish.

The runtime steps associated with asynchronous replication are shown in
Figure~\ref{fig:async_rep}.

\begin{figure}
{\scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
TX_BEGIN {
  Execute RPC call and modify local NVM
  while(!replication_complete);
  if(replication_failed) {
    TX_ABORT;
  }
} TX_END
\end{verbatim}
}
\caption{Asynchronous Replication}
\label{fig:async_rep}
\end{figure}

\subsection{Synchronous Replication}

\section{Multi-node Programming Model}

\subsection{Example: Distributed Transactions}

>>>>

This paper presents Cyclone - fault tolerance middleware that exposes a
replicated and durable heap. This heap can be directly manipulated using loads and
stores. Cyclone provides both failure atomicity as well as
strongly consistent replication when updates to the heap are enclosed in
software transactions thereby giving strong guarantees to programmers
with minimal effort on their part. Under the hood Cyclone decouples modification
of local persistent memory from replication to maximize performance. In
addition, unlike traditional systems that provide strong replication, Cyclone
does require programs to be provisioned for checkpoints since NVM is directly
manipulated by programmers and therefore up to date till the last executed
transaction. Instead, Cyclone provides a novel NVM copy mechanism to bring up
new nodes added to the cluster.

\end{abstract}

@@

\begin{figure}
{ \scriptsize
\begin{verbatim}

@@

\section{Failure Atomicity}
Programmers typically find it hard to reason about failures when writing
code. This is a problem when directly manipulating data structures in persistent
memory. Depending on the specifics of the system architecture, updates to
persistent memory can be persisted out of order thereby leaving the data
structure in an inconsistent state. To illustrate this consider a failure during
execution of the code to add a node to a linked list in
Figure~\ref{fig:example}. If either of the two updates to the next fields of the
previous and new node fail to be persisted, the linked list is left in a
disconnected state with one or more connected nodes inaccessible.

@@

\begin{figure}
{ \scriptsize
\begin{verbatim}
TX_BEGIN()
{
  Initialize undo log
}

@@

A crucial feature of Cyclone is that it abstracts the details of the persistence
mechanism away from programmers. This allows the same code to be deployed on
clusters with a persistent storage medium such as 3D XPoint or on clusters with
battery backed NVDIMMs for example. In the former case, explicit {\tt PCOMMIT}
instructions are necessary for persistence [ref PMFS paper], while in the latter
case no such instructions are necessary as on failure the CPU caches and DRAM
are flushed to stable storage under battery (or capacitor) power. Even in the
NVDIMM case however, the undo log can be desirable to protect against OS or
application crashes to ensure that the persistent memory is always left in a
consistent state. Cyclone's programmer visible features are designed to enable
the construction of applications that work regardless of the underlying
persistence mechancisms.

\section{Replication}
Cyclone replicates the state of the NVM heap across multiple machines in a
cluster. Cyclone provides strong consistency - a majority of machines in the
quorum of replicas always agrees on the current state of the NVM heap. Unlike
traditional approaches to strong replication however, Cyclone is designed to
decouple replication from execution of user code modifying NVM. It does so by
encapsulating changes to the NVM heap in RPC calls driven by client requests.
We chose this model since it closely reflects the use cases most suitable for
NVM-based clusters.

Cyclone replicates state \emph{by executing the same sequence of RPC calls on
  all members of the quorum}. By executing the same sequence of RPC calls on all
machines we end up with the same state\footnote{We discuss how to handle
  non-determinism in Section~\ref{sec:synch}}. Cyclone uses RAFT [cite usenix
  raft paper] under the hood for consensus on the contents of the log of RPC
calls - although any other similar state machine replication protocol such as
Paxos is also suitable. Cyclone provides the means to achieve exactly once
semantics to clients by numbering RPC calls and remembering the result of the
last executed RPC call.

@@

\subsection{Asynchronous Replication}
By default, Cyclone decouples execution of the RPC call on any replica from its
replication. This is particularly beneficial to performance since it allows
overlapping the update of non-volatile state with the network I/O associated
with replication. Persisting RAFT log entries, making the upcall to the
application and executing the RPC call in a failure atomic transaction takes a
significant amount of time and therefore it makes sense to overlap this with the
round-trip delay on the network for replication. This is possible in Cyclone
since we use failure atomic transactions that can be rolled back. Replication
can fail due to various reasons such as failure of nodes, a network partition or
a new leader getting elected in the underlying replication protocol
(RAFT). Since failure atomic transactions can be rolled back, we are able to
speculatively apply the changes to NVM without waiting for replication to
finish.

The runtime steps for asynchronous replication are shown in
Figure~\ref{fig:async_rep}.

\begin{figure}
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
TX_BEGIN {
  Execute RPC call and modify local NVM
  // Block till replication result known
  if(replication failed) {
    TX_ABORT;
  }
} TX_END
\end{verbatim}
}
\caption{Asynchronous Replication}
\label{fig:async_rep}
\end{figure}

\subsection{Synchronous Replication}
\label{sec:synch}
Asynchronous replication provides good performance but lacks guarantees
occasionally necessary for distributed systems paradigms such as output commit and
the ability to integrate code that has non-deterministic execution paths or side
effects - such as reading the time. Cyclone therefore provides programmers the
ability to closely synchronize execution across the replicas using synchronous
replication. Synchronous replication extends the programming model provided by
cyclone - but neccessitates some awareness on the part of the user about
the difference between the leader and follower replicas. We show later
(Sec~\ref{sec:examples}) how this awareness enables the crafting of efficient
fault tolerant applications with little effort on the part of the programmer.

Figure~\ref{fig:sync_rep} shows how synchronous replication operates.  Execution
of the RPC call \emph{waits until the call and its arguments have been
  replicated.} This crucial difference from asynchronous RPC execution ensures
that any externally visible side effects of the RPC such as calls to other
machines only happen after it is guaranteed that the effects of the RPC call
will be committed to the replicas. This provides the output commit property
necessary for some fault tolerance features in applications. Next the RPC call
is executed on the leader and the result of the RPC call is replicated to the
followers. The followers then proceed to execute the RPC call locally taking
into account the result from the leader. This result can include data to
synchronize non-deterministic steps such as reading the time on the replica.

This process is made conditional on the view (current configuration of
replicas and choice of leader) staying unchanged. To successfully complete
execution, the leader must be able to execute the RPC call and replicate the
result within the current view. This ensures that in the event of a network
partition and later repair an old execution of a previous leader cannot commit a
log record with results from a synchronously executed RPC call.

\begin{figure*}
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
while(true) {
  TX_BEGIN {
    // Block till replication result known
    if(replication failed) {
      TX_ABORT;
    }  
    if(is_leader && !have_execution_result) {
      Execute RPC call and modify local NVM
      replicate rpc call result in current view;
      // Block till replication result known
      if(result replication failed) {
        TX_ABORT;
      }
    }
    else {
      // Block till replication done or view changes
      if(!result replication complete) {
        TX_ABORT;
      }
      Execute RPC call to modify local NVM with result from leader
    }
  }
} TX_END
\end{verbatim}
}
\caption{Synchronous Replication}
\label{fig:sync_rep}
\end{figure*}

\section{Example Applications}
\label{sec:examples}
